How residence time constrains diversity

### 2015-06-20 paper/Manuscript.Rmd

*Simulated life history-*
The community was sampled at random *N* times with replacement for each simulated time unit. This theoretically gives every individual the opportunity to grow and reproduce. Sampled individuals consume resources according to species specific rates of uptake, if there is a resource present at the individual’s location. Uptake increases the individual cell quota and decreases ambient resources within the immediate location of the individual. The individual’s quota is then decreased according to a species specific physiological maintenance cost. Individuals then reproduce with a probability equal to their cell quota and results in an equal  division of the cell quota between the individual and its daughter. The daughter is given a unique individual ID which is appended to the species ID of its mother, to maintain heritage. Sampled individuals die if their cell quota is less than or equal to 0. Individuals and resource particles flow out when they pass the downstream edge of the environment.

Simulated environmental complexity—We constructed the model to capture environmental heterogeneity and the aggregation of resources and individuals. The model was parameterized with an arbitrary environmental gradient that varied regularly across a randomly chosen spatial axis in a randomly chosen direction. Barriers were constructed within the environment to prevent unhindered flow and to generate aggregation and realistic hydrological flow.

Simulated species differences—To allow species differences in physiological traits and environmental optima, we allowed species-specific values for maximum rate of uptake rate, environmental optima, and maintenance costs to be normally distributed random variables where both the means and the standard deviations of the sampling distributions were chosen at random at the start of the simulations. Whenever an individual of a new species entered the local community, values for maximum uptake rate, environmental optima, physiological maintenance cost, and minimum cell quota were chosen at random according to the respective normal distributions for those variables.

Evaluating model performance in real-time—The model simulates over thousands of generations until a state of stationarity is reached. Because increasing values of $\tau$ leads to slow population dynamics and because the simulations can vary over three orders of magnitude in $\tau$, we designed the model to track its own performance to ensure that time points gathered from the model were not taken during ‘burn in’ and to ensure time points were not auto-correlated. For each generation (i.e. a number of time steps equal to N, which is not constant) after the first 1000 generations, the model evaluates whether it has reached stationarity by calculating the value of the Hurst exponent (H) and the p-value of the variance ratio test for values of N recorded from the last 500 generations. Stationarity is assumed to be reached when H < 0.5 and p < 0.05 (cite). All time points before this ‘burn-in’ period are discarded. After reaching stationarity, the model runs for another 1000 generations and then stops to check if there are at least 100 sufficiently non-autocorrelated time points, as determined by the time lag after which absolute values of the autocorrelation function are less than 0.1. If not, the model runs for another 1000 generations, repeating the process until 100 non-autocorrelated time points is obtained.

Simulating over residence time ($\tau$)—We ran the model to quasi-equilibrium for each value of $\tau$ in a series of values spanning more than two orders of magnitude, starting with $\tau$ = 0.001 and then doubling until $\tau$ = 0.512. The model was run with the same combination of resource, community ecological, and environmental (thermal, physicochemical, physical complexity) values for each value of $\tau$ in the series. After running over all $\tau$ values, the model was re-parameterized at random with a new combination of resource parameters (size, number, and types of resource particles), habitat complexity parameters (size, number, and types of physical barriers), chemical environmental parameters, thermal parameters (average temperature, direction and severity of temperature gradient), and average community-level values (average specific values for uptake rate, maintenance, individual size, motility, attachment). See table 1.

	Because the model is individual-based, information intensive, physiologically realistic, ecologically complex, fluid dynamic, self-evaluating, it cannot be run over arbitrarily slow or fast residence times. During each simulation, more than a gigabyte of information is held in memory. This is unlike analytically tractable and computationally simple Markov models or closed-form analytical models, which are more useful for obtaining precise quantitative predictions according to a single theoretical framework via traditional mathematical tools, but are less useful for modeling highly complex multivariate problems and for examining how individual-level interactions lead to aggregate behaviors. In our case, we have integrated components from metabolic theory and ecological scaling, neutral and niche theory, metacommunity theory, fluid dynamics, chemostat theory, and life history theory into a self-evaluating model that tracks the life history, location, and heritage of millions of individuals.

>>>>

*Simulated life history-*The community was sampled at random *N* times with replacement for each simulated time unit. This theoretically gives every individual the opportunity to grow and reproduce. Sampled individuals consume resources according to species specific rates of uptake, if there is a resource present at the individual’s location. Uptake increases the individual cell quota and decreases ambient resources within the immediate location of the individual. The individual’s quota is then decreased according to a species specific physiological maintenance cost. Individuals then reproduce with a probability equal to their cell quota and results in an equal  division of the cell quota between the individual and its daughter. The daughter is given a unique individual ID which is appended to the species ID of its mother, to maintain heritage. Sampled individuals die if their cell quota is less than or equal to 0. Individuals and resource particles flow out when they pass the downstream edge of the environment.

*Simulating environmental complexity—*We constructed the model to capture environmental heterogeneity and the aggregation of resources and individuals. The model was parameterized with an arbitrary environmental gradient that varied regularly across a randomly chosen spatial axis in a randomly chosen direction. Barriers were constructed within the environment to prevent unhindered flow and to generate aggregation and realistic hydrological flow.

*Simulating species differences*—To allow species differences in physiological traits and environmental optima, we allowed species-specific values for maximum rate of uptake rate, environmental optima, and maintenance costs to be normally distributed random variables where both the means and the standard deviations of the sampling distributions were chosen at random at the start of the simulations. Whenever an individual of a new species entered the local community, values for maximum uptake rate, environmental optima, physiological maintenance cost, and minimum cell quota were chosen at random according to the respective normal distributions for those variables.

*Simulating over residence time ($\tau$)—*We ran the model to quasi-equilibrium for each value of $\tau$ in a series of values spanning more than two orders of magnitude, starting with $\tau$ = 0.001 and then doubling until $\tau$ = 0.512. The model was run with the same combination of resource, community ecological, and environmental (thermal, physicochemical, physical complexity) values for each value of $\tau$ in the series. After running over all $\tau$ values, the model was re-parameterized at random with a new combination of resource parameters (size, number, and types of resource particles), habitat complexity parameters (size, number, and types of physical barriers), chemical environmental parameters, thermal parameters (average temperature, direction and severity of temperature gradient), and average community-level values (average specific values for uptake rate, maintenance, individual size, motility, attachment). See table 1.

Because the model is individual-based, information intensive, physiologically realistic, ecologically complex, fluid dynamic, it cannot be run over arbitrarily slow or fast residence times. During each simulation, more than a gigabyte of information is held in memory. This is unlike analytically tractable and computationally simple Markov models or closed-form analytical models, which are more useful for obtaining precise quantitative predictions according to a single theoretical framework via traditional mathematical tools, but are less useful for modeling highly complex multivariate problems and for examining how individual-level interactions lead to aggregate behaviors. In our case, we have integrated components from metabolic theory and ecological scaling, neutral and niche theory, metacommunity theory, fluid dynamics, chemostat theory, and life history theory into a self-evaluating model that tracks the life history, location, and heritage of millions of individuals.

## Results

*Autocorrelated residence times and right-skewed residence time distributions——*As expected, residence times of individuals, resource particles, inert tracers, and most any non-permanent component of the system were highly correlated. Right-skewed distributions characterized the life history of individuals, the presence of resource particles, and the tenure of species and traits (Figure ?). Residence time distributions were generally well approximated by lognormal and power-law models (e.g. Pareto, Zipf, Zipf-Mandelbrot).
